---
title: "Premier League Blog Post"
author: "Cooper Olney"
date: "2026-02-04"
categories: [analysis]
---

# Introduction
The dataset used in this analysis contains English Premier League match data from the 2021-2022 season. The dataset includes 380 observations, corresponding to the 380 matches played over the course of a full season. Each match records performance statistics for both the home and away teams.

The variables I am most interested in are shots taken and fouls committed by the home and away teams. These variables provide insight into team playing styles, particularly whether home teams tend to play a more attacking style by generating more shots, and whether away teams are forced into a more defensive approach, potentially committing more fouls. Other metrics to also consider are corner kicks and yellow cards, as teams typically win more corners the more they attack, while yellow cards can indicate defensive play or time wasting tactics.

This leads to my primary question of interest:

**Do home teams play a more attacking style of soccer, while away teams adopt a more defensive style of play?**

This data was accessed through the TidyTuesday project on a GitHub repository. The original source of the data is from Evan Gower on Kaggle.

Some additional context about the Premier League is important for interpreting this data. The league consists of 20 teams, with each team playing 38 matches, once at home and once away against every other team. Match outcomes are scored as 3 points for a win, 1 point for a draw, and 0 points for a loss. Playing at home is widely believed to provide a competitive advantage due to factors such as crowd support, less traveling, better stadium amenities, and even differences in pitch dimensions.

As a result, teams often approach away matches more conservatively. Securing a draw away from home is typically considered a successful outcome, which encourages defensive strategies. In contrast, home matches are seen as opportunities to earn the maximum number of points, leading teams to adopt a more attacking style of play. Away teams may also attempt to slow the pace of the game through fouling or time-wasting, which can be reflected in higher foul counts and yellow cards.

Overall, these contextual factors motivate an exploration of whether match statistics such as shots taken and fouls committed support the commonly held belief that home teams attack more, while away teams defend more.


# Primary Visualizations

## Setting up the Data
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r message=FALSE, warning=FALSE}
soccer <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-04-04/soccer21-22.csv')
```

```{r message=FALSE, warning=FALSE}
soccer <- soccer |> mutate(shot_diff = HS - AS,
                 foul_diff = HF - AF,
                 corner_diff = HC - AC)
```

## Plot 1: Difference in Shots: Home vs Away
```{r message=FALSE, warning=FALSE}
ggplot(data = soccer, aes(x = shot_diff)) +
  geom_histogram(binwidth = 4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs( title = "Difference in Shots: Home vs Away", 
        x = "Home Shots − Away Shots", 
        y = "Number of Matches") +
  xlim(-29,29) +
  theme_minimal()
```

The distribution of the histogram is approximately normal, with its peak occurring when the home teams holds a 2-6 shot advantage. Looking at extreme shot differences, the home team out shooting the away team by 15 or more shots is more common than the other way around. This pattern suggests that home teams are generally more likely to play an attacking style of soccer. However, there is still a presence of negative values showing that matches end with the away team taking more shots. This indicates the home field advantage alone does not fully determine a team's style of play, and that other factors such as overall team quality likely plays a role. 

## Plot 2: Difference in Fouls: Home vs Away
```{r message=FALSE, warning=FALSE}
ggplot(data = soccer, aes(x = foul_diff)) +
  geom_histogram(binwidth = 3) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs( title = "Difference in Fouls: Home vs Away", 
        x = "Home Fouls − Away Fouls", 
        y = "Number of Matches") + 
  xlim(-20, 20) +
  theme_minimal()
```
The histogram peaks around a shot difference of 0, showing that in most matches there is little to no difference in the number of fouls committed by the home and away teams. However, the distribution is skewed to the left, with several negative outliers where the away team has committed many more fouls. This suggests that in certain matches, away teams are forced into a more defensive style of play, relying heavily on fouling to disrupt the home team's attacks and slow the pace of the game.

## Plot 3: Shot Difference by Match Outcome
```{r message=FALSE, warning=FALSE}
ggplot(data = soccer, aes(x = FTR, y = shot_diff, fill = FTR)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed") + 
  scale_x_discrete(labels = c("A" = "Away Win", "D" = "Draw", "H" = "Home Win")) +
  labs( title = "Shot Difference by Match Outcome", 
        x = "Final Match Result", 
        y = "Home Shots − Away Shots") + 
  theme_minimal() + 
  ylim(-29, 29) +
  theme(legend.position = "none") +
  coord_flip()
```
The boxplots show a clear relationship between shot difference and the final match result. When the away team wins, the median shot difference is around -2, meaning that away victories typically occur when the away team is able to generate slightly more attacking opportunities than the home team.

For drawn matches, the median shot difference is about 2 suggesting that home teams still tend to take more shots even when the match ends level. Which is consistent with the idea that away teams often play more defensively and are satisfied with drawing matches away from home. When the home team wins, the median show difference rises to around 7, showing that home victories are strongly associated with attacking play style where the home team dominates the possession of the ball.

Overall, this plot supports the narrative that home teams generally play a more attacking style, while the away team adopts more defensive strategies. The plot also shows that the amount of attacking pressure a team can apply is potentially associated with the outcome of the match.


# Conclusion
The analysis provides evidence that home teams tend to play a more attacking style and away teams adopt a more defensive approach, there are several limitations to consider in the analysis. Variables such as shots and fouls does not capture exactly how a team plays and all of the tactical decisions made during a match. Style of play can change based on differences in team quality, player injuries, or time in between the most recent match can all strongly influence how a team plays in the current match.

With additional data, a next step could be to incorporate team strength into the analysis. Teams could be grouped into upper, middle and lower tier categories based on where they finished in the Premier League table to find out how home field advantage interacts with overall talent. This could provide a deeper understanding of how playing style changes depending on venue. My hypothesis would be that middle tier teams would show the greatest change in style when switching between home and away matches.

# Connection to Class Ideas
For the first two plots, I chose histograms because they effectively display a single quantitative variable, showing the center, shape, and spread of the data, as well as roughly how many observations there are in total. Since I was primarily interested in the distribution of the shot and foul differences, histograms made a lot of sense. The side-by-side boxplot works well for comparing a quantitative variable across a categorical variable, making it easy to see differences in distributions. I added color simply to make the plot visually appealing with an unordered color palette because a sequential or diverging color palette wasn't needed in this case.

The plots are intentionally simple, so viewers can quickly understand the patterns without being overwhelmed. The dashed vertical lines highlight that positive differences correspond to the home team having more shots or fouls, and negative differences indicate the away team had more. I also set symmetric x-axis limits around zero to prevent skewed or inaccurate representations of the data, and to emphasize that the plot is split between home and away advantage. This ensures that zero remains central, reinforcing the narrative the half the plot represents a home advantage and the other half represents an away advantage.

